---
title: "Notes muscle mass and strength"
output: html_document
---

# Results
```{r, include=FALSE, warning=FALSE, message=FALSE}

# This is the code I used to calculate the relationship between lean body mass and 3RM squat at T1.

######## Preparing for the tests ######## 
# Loading needed packages
library(tidyverse)
library(knitr)
library(broom)
library(readxl)


# Downloading and saving the file
download.file(url = "https://ndownloader.figstatic.com/files/14702420", destfile = "./data/hypertrophy.csv")

# Reading and attaching file to "hypertrophy"
hypertrophy <- read_csv("./data/hypertrophy.csv")

# Selecting the variables of interest (DXA_LBM_1, Squat_3RM_kg)
hyp <- hypertrophy %>% # Attaching to "hyp"
  select(SUB_ID, DXA_LBM_1, Squat_3RM_kg, ) %>% # Selecting variables of interest
  
  filter(!is.na(Squat_3RM_kg), # Removing missing values in Squat_3RM_kg
         !is.na(DXA_LBM_1))  # Removing missing values in DXA_LBM_1


######## Correlations ######## 
# Testing the correlation between "DXA_LBM_1" and "Squat_3RM_kg" using "hyp" as data
corr <- cor(hyp$DXA_LBM_1, hyp$Squat_3RM_kg)

c <- cor.test(hyp$DXA_LBM_1, hyp$Squat_3RM_kg) %>% print()
# The correlation is 0.6741231. Perfect correlation = -1 or 1, no correlation = 0
# Ergo, in this data there is a good correlation between muscle size and strength



######## Storing the statistics as objects to use in the report ######## 

# The p-value
p <- c$p.value  # p = (4.421018e-05)

# The correlation = 0.6741231 (see )
corr

# The test statistic
t <- c$statistic  # t = 4.8295

# Confidence interval
ci1 <- c$conf.int[1]   # Lower CI = 0.4145287
ci2 <- c$conf.int[2]   # Higher CI = 0.8322639, conf level = 0.95

# Degrees of freedom
df <- c$parameter  # df = 28



######## Regression model ######## 
# Using a linear model (lm()) to fit the data
m1 <- lm(Squat_3RM_kg ~ DXA_LBM_1, data = hyp) # Storing the regression model as m1
summary(m1) # Showing the results

# Plotting the regression model
rmodel <- hyp %>%
  ggplot(aes(DXA_LBM_1, Squat_3RM_kg)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Lean body mass (kg)",
       y = "3RM squat (kg)",
       title = "Figure1: The relationship between muscle mass and strength") +
  theme_minimal() %>% print()


######## Checking for constant variance ######## 
# Creating a residual plot, first getting the residuals and fitted values
hyp$resid <- residuals(m1)  # Residuals
hyp$fitted <- fitted(m1)  # Fitted values

# Plotting the residuals against the fitted values
hyp %>%
  mutate(st.resid = resid/sd(resid)) %>% # Making standardized residuals
  ggplot(aes(fitted, st.resid)) + 
  geom_hline(yintercept = 0) +
  geom_point(size = 3, fill = "lightblue", shape = 21) +
  labs(x = "Fitted values",
       y = "Standardized residuals") +
  theme_minimal()
    # All the residuals, except for one, is less than 2.0 SD away from their predicted value. The variance is quite constant.



######## Checking for normal residuals ########
# Checking if the residuals are normal by creating a plot that plot every observed residual against its theoretical position in a normal distribution
hyp %>%
  mutate(st.resid = resid/sd(resid)) %>% 
  ggplot(aes(sample = st.resid)) +
  stat_qq(size = 3, fill = "lightblue", shape = 21) +
  stat_qq_line() +
  labs(x = "Theoretical",
       y = "Sample") +
  theme_minimal()
  # Most of the values follows the straight line. The value to the left lies further away from the line than the others. Still, the data is quite normal distributed. 

# Can also show the above using residual plots (four plots).
plot(m1)



######## Independent observations ########
# The data comes from independent and paired observations, at T1.



######## Do problematic observations matter? ########
# Saw that there was one potentially problematic observation in the residual plot. 
# Labeling the observations in the residual plot to find out what observation is problematic.

hyp %>%
  mutate(st.resid = resid/sd(resid)) %>%
  ggplot(aes(fitted, st.resid, label = SUB_ID)) + 
  geom_hline(yintercept = 0) +
  geom_point(size = 3, fill = "lightblue", shape = 21) +
  geom_label(nudge_x = 20, nudge_y = 0) +
  labs(x = "Fitted values",
       y = "Standardized residuals") +
  theme_minimal()
# Finding that it is subject MRV019 that has the largest residual and and potentially could be a problematic observation

# Checking if the potentially problematic observation MRV019 really is problematic. Doing the model without the observation to see if this changes the conclusion of the analysis.
hyp_reduced <- hypertrophy %>% # Creating a new object without MRV016
  select(SUB_ID, DXA_LBM_1, Squat_3RM_kg) %>% # Selecting variables of interest
  filter(SUB_ID != "MRV019") %>% # Removing observation MRV019
  filter(!is.na(Squat_3RM_kg), # Removing missing values in Squat_3RM_kg
         !is.na(DXA_LBM_1))  # Removing missing values in DXA_LBM_1


# Making a regression model without MRV019
m1_reduced <- lm(Squat_3RM_kg ~ DXA_LBM_1, data = hyp_reduced)
summary(m1_reduced)
# All the values change a bit

# The delta beta calculates the percentage change in the slope as a consequence of removing the observation with the greatest residual
delta_beta <- 100 * (coef(m1_reduced)[2]/coef(m1)[2] - 1) 
# Removing the outlier changes the slope with -0.05035753 %. Ergo, not a lot.

# We can conclude that the potentially problematic observation do not matter, and that we can safely use all the data points in the analysis, as done above under "correlation" and "regression model"

# We can also conclude that the correlation and the regression model provides the same results


```

```{r, inklude=FALSE, results='asis', warning=FALSE, message=FALSE}
rmodel

# Modellen til lean body mass
```

There was a `r round(corr, 2)` correlation between lean body mass and 3RM squat at T1 (95% CI: [`r round(ci1, 2)`, `r round(ci2, 2)`], *t*(`r round(df, 1)`) = `r round(t,3)`, *p* = `r round(p, 2)`).


## Discussion
The correlation estimate (0.67) is between 0.5 and 0.7, witch indicates a *large* positive correlation between lean body mass and 3RM squat at T1 (Hopkins et al., 2009). This is supported by the *p*-value (`r round(p, 2)`).

As the confidence interval do not contain 0 (95% CI: [`r round(ci1, 2)`, `r round(ci2, 2)`]), we can with some certainty reject the null hypothesis that states that "at T1, there is no correlation between muscle mass measured as lean body mass and muscle strength measured as 3RM squat."

The findings from Haun et al. (2019) therefore suggests that there is a *large* correlation between muscle mass and strength, measured as lean body mass and 3RM squat at T1.
