---
title: "Relationship between muscle mass and strength"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Introduction
Haun et al. (2019) conducted their study on 30 trained men who performed a 6 weeks resistance training intervention. Two groups were formed based on training outcome measures, where the top and bottom third of the data set were assigned to a high- (*n* = 10) and low-responder group (*n* = 10), respectively. The goal of the study was to examine if HIGH responders would experience greater increases in myonuclear accretion as well as biomarkers related to ribosome biogenesis, androgen signaling, mTORc1 signaling, and mitochondrial biogenesis relative to LOW responders.

The goal of this report is to estimate the relationship between muscle mass and strength, measured as lean body mass and 3RM squat at T1, respectively. The whole data set was used, meaning we used data from all the participants.

#### Null hypothesis
> At T1, there is no correlation between muscle mass measured as lean body mass and muscle strength measured as 3RM squat

#### Alternative hypothesis
> At T1, there is a positive correlation between muscle mass measured as lean body mass and muscle strength measured as 3RM squat

## Methods
Pearson's product-moment correlation was used to estimate the correlation between lean body mass and 3RM squat at T1. r < 0,1 = trivial, 0,1-0,3 = small, 0,3-0,5 = moderate, 0,5-0,7 = large, 0,7-0,9 = very large, 0,9 = almost perfect and 1.0 = perfect correlation (Hopkins et al., 2009). The data was plotted as single data points, and a straight line representing the linear regression model was added.

## Results
```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
######## Preparing for the tests ######## 
# Loading needed packages
library(tidyverse)
library(knitr)
library(broom)

# Downloading and saving the file
download.file(url = "https://ndownloader.figstatic.com/files/14702420", destfile = "./data/hypertrophy.csv")

# Reading and attaching file to "hypertrophy"
hypertrophy <- read_csv("./data/hypertrophy.csv")

# Selecting the variables of interest (DXA_LBM_1, Squat_3RM_kg)
hyp <- hypertrophy %>% # Attaching to "hyp"
  select(SUB_ID, DXA_LBM_1, Squat_3RM_kg, ) %>% # Selecting variables of interest
  
  filter(!is.na(Squat_3RM_kg), # Removing missing values in Squat_3RM_kg
         !is.na(DXA_LBM_1))  # Removing missing values in DXA_LBM_1



######## Correlations ######## 
# Testing the correlation between "DXA_LBM_1" and "Squat_3RM_kg" using "hyp" as data
c <- cor.test(hyp$DXA_LBM_1, hyp$Squat_3RM_kg)
# The correlation is 0.6741231. Perfect correlation = -1 or 1, no correlation = 0
# Ergo, in this data there is a good correlation between muscle size and strength



# Plotting the regression model
hyp %>%
  ggplot(aes(DXA_LBM_1, Squat_3RM_kg)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Lean body mass (kg)",
       y = "3RM squat (kg)",
       title = "Figure1: The relationship between muscle mass and strength") +
  theme_minimal()




######## Storing the statistics as objects to use in the report ######## 

# The p-value
p <- c$p.value  # p = (4.421018e-05)

# The correlation = 0.6741231 (see )

# The test statistic
t <- c$statistic  # t = 4.8295

# Confidence interval
ci1 <- c$conf.int[1]   # Lower CI = 0.4145287
ci2 <- c$conf.int[2]   # Higher CI = 0.8322639, conf level = 0,95

# Degres of freedom
df <- c$parameter  # df = 28


```

There was a 0.67 correlation between lean body mass and 3RM squat at T1 (95% CI: [`r round(ci1, 2)`, `r round(ci2, 2)`], *t*(`r round(df, 1)`) = `r round(t,3)`, *p* = `r round(p, 2)`).


## Discussion
The correlation estimate (0.67) is between 0.5 and 0.7, witch indicates a *large* positive correlation between lean body mass and 3RM squat at T1 (Hopkins et al., 2009). This is supported by the *p*-value (`r round(p, 2)`).

As the confidence interval do not contain 0 (95% CI: [`r round(ci1, 2)`, `r round(ci2, 2)`]), we can with some certainty reject the null hypothesis that states that "at T1, there is no correlation between muscle mass measured as lean body mass and muscle strength measured as 3RM squat."

The findings from Haun et al. (2019) therefore suggests that there is a *large* correlation between muscle mass and strength, measured as lean body mass and 3RM squat at T1.


## References
Haun, Cody T., Christopher G. Vann, C. Brooks Mobley, Shelby C. Osburn, Petey W. Mumford, Paul A. Roberson, Matthew A. Romero, et al. 2019. “Pre-Training Skeletal Muscle Fiber Size and Predominant Fiber Type Best Predict Hypertrophic Responses to 6 Weeks of Resistance Training in Previously Trained Young Men.” Journal Article. Frontiers in Physiology 10 (297). https://doi.org/10.3389/fphys.2019.00297.
Hopkins, W. G., Marshall, S. W., Batterham, A. M. & Hanin, J. (2009). Progressive statistics for studies in sports medicine and exercise science. Med Sci Sports Exerc, 41(1), 3-13. https://doi.org/10.1249/MSS.0b013e31818cb278

```{r, include=FALSE}
#      Er dette nødvendig å ha med?
# _____________________________________________ #
###### Trengs egentlig regression model? Holder det med correlation og en figur?

######## Regression model ######## 
# Using a linear model (lm()) to fit the data
rm <- lm(Squat_3RM_kg ~ DXA_LBM_1, data = hyp) # Storing the regression model as m1
summary(rm) # Showing the results



# Printing the regression table using the broom package (tidy()) and kable()
tidy(rm) %>% 
  kable(col.names = c("", "Estimate", "SE", "t-statistic", "p-value"), # Column names
              digits = c(NA, 1, 1, 2, 4)) # Deciding how many decimals at each value

# _____________________________________________ #





######## Comparing regression model and correlation ######## 
# Display the p-value for the regression coefficient
coef(summary(rm))[2, 4]       # Ta bort??

# Display the p-value for the correlation coefficient
p <- c$p.value  # The p-values are the same (4.421018e-05)
```


