---
title: "Relationship between muscle mass and strength"
output: html_document
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

# The relationship between average cross-sectional area and 3 RM squat

## Introduction
Haun et al. [-@Haun2019] conducted their study on 30 trained men who performed a 6 weeks resistance training intervention. Two groups were formed based on training outcome measures, where the top and bottom third of the data set were assigned to a high- (*n* = 10) and low-responder group (*n* = 10), respectively. The goal of the study was to examine if high responders would experience greater increases in myonuclear accretion as well as biomarkers related to ribosome biogenesis, androgen signaling, mTORc1 signaling, and mitochondrial biogenesis relative to low responders.

Maughan et al. [-@Maughan1983] have found a positive correlation between muscle strength and cross-sectional area in the knee-extensor muscles in both males and females. The goal of this report is to estimate the relationship between muscle mass and strength, measured as average cross-sectional area and 3 RM squat, respectively, both at T1. The whole data set was used, meaning we used data from all of the participants.

#### Null hypothesis
> At T1, there is no correlation between muscle mass measured as average cross-sectional area and muscle strength measured as 3 RM squat, both when and when not controlling for body mass

#### Alternative hypothesis
> At T1, there is a positive correlation between muscle mass measured as average cross-sectional area and muscle strength measured as 3 RM squat, both when and when not controlling for body mass

## Methods
Pearson's product-moment correlation was used to estimate the correlation between average cross-sectional area and 3 RM squat at T1. r < 0.1 = trivial, 0.1-0.3 = small, 0.3-0.5 = moderate, 0.5-0.7 = large, 0.7-0.9 = very large, 0.9 = almost perfect and 1.0 = perfect correlation [@Hopkins2009]. A regression model was also used to fit the data, with the average cross-sectional area as the predictor, and 3 RM squat as the dependent variable. The regression model also controlled for body mass. Tests to check for constant variance, normal residuals and problematic observations were performed. No problematic observations were found, so we could safely use all the observations in the analysis

## Results

```{r, include=FALSE, warning=FALSE, message=FALSE}
# I use this chunck to run all the code


######## Preparing for the tests ######## 
# Loading needed packages
library(tidyverse)
library(knitr)
library(broom)
library(readxl)

# Downloading and saving the file
download.file(url = "https://ndownloader.figstatic.com/files/14702420", destfile = "./data/hypertrophy.csv")

# Reading and attaching file to "hypertrophy"
hypertrophy <- read_csv("./data/hypertrophy.csv") 

# Selecting the variables of interest (AVG_CSA_T1, Squat_3RM_kg)
hyp <- hypertrophy %>% # Attaching to "hyp"
  select(SUB_ID, AVG_CSA_T1, Squat_3RM_kg, T1_BODY_MASS) %>% # Selecting variables of interest
  filter(!is.na(Squat_3RM_kg), # Removing missing values in Squat_3RM_kg
         !is.na(AVG_CSA_T1))  # Removing missing values in AVG_CSA_T1

# mutate(Squat_3RM_kg = (Squat_3RM_kg/1000)) %>%

#----------------------------------------------------------------------------
# Can use correlation to see if there is a relatonship 

######## Correlations ######## 
# Testing the correlation between "AVG_CSA_T1" and "Squat_3RM_kg" using "hyp" as data
corr <- cor(hyp$AVG_CSA_T1, hyp$Squat_3RM_kg)

c <- cor.test(hyp$AVG_CSA_T1, hyp$Squat_3RM_kg)  

c
# Lage dette som tabell? Se måte å gjøre det på under "regression model"
# The correlation is 02068658. Perfect correlation = -1 or 1, no correlation = 0
# Ergo, in this data there is no correlation between muscle size and strength


######## Storing the statistics as objects to use in the report ######## 

# The p-value
p <- c$p.value   # p = 0.9135975

# The correlation
corr # corr = 02068658

# The test statistic
t <- c$statistic  # t = 0.1094865 

# Confidence interval
ci1 <- c$conf.int[1]    # Lower CI = -0.3421325
ci2 <- c$conf.int[2]   # Higher CI = 0.3781376, conf level = 0.95

# Degrees of freedom
df <- c$parameter # df = 28



#----------------------------------------------------------------------------
# Can make a better model when adding body mass and using a regression model

######## Regression model ######## 
# Using a linear model (lm()) to fit the data
m1 <- lm(Squat_3RM_kg ~ AVG_CSA_T1 + T1_BODY_MASS, data = hyp) # Storing the regression model as m1
summary(m1) # Showing the results. *Bruk tallene fra linje 2?*

# Printing the regression table using the broom package (tidy()) and kable()
rtable <- tidy(m1) %>% 
  kable(col.names = c("", "Estimate", "SE", "t-statistic", "p-value"), # Column names
        digits = c(NA, 4, 4, 4, 4),
        caption = "Table1: The relationship between muscle mass and strength. Using a regression model with the average cross-sectional area as the predictor, and 3 RM squat as the dependent variable") # Deciding how many decimals at each value


rtable
# Loock at line 2
# For every unit increase in CSA, 3 RM sqaut increases with 0,0028 kg.

# Plotting the regression model
rmodel <- hyp %>%
  ggplot(aes(AVG_CSA_T1, Squat_3RM_kg)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Average CSA",
       y = "3 RM squat (kg)") +
  theme_minimal()

rmodel


# Coefficients
coef(m1)[1] # 25.65533 
coef(m1)[2] # 0.002812117  Controlled for body mass
coef(m1)[3] # 1.137279 

# Confidence intevals
# Linje 1
confint(m1)[1] # -28.75575
confint(m1)[4] # 80.06642
# Linje 2 (disse som skal brukes)
confint(m1)[2] # -0.004047546 (low CI)
confint(m1)[5] # 0.00967178 (high CI)
# Linje 3
confint(m1)[3] # 0.6264106
confint(m1)[6] # 1.648148

# Wanted to connect p-value, df osv to objects so I could use them in the text in the discussion, but did not now how

## ---------------------------------------------------------------------------------
# Tests to check for constant variance, normal residuals and problematic observations


# Did not have to do this, but as I did, I am just saving the code here for later use
# In the text I will write that the testes were performed, and no problematic observations were found, just like you said I could. 



######## Checking for constant variance ######## 
# Creating a residual plot, first getting the residuals and fitted values
hyp$resid <- residuals(m1)  # Residuals
hyp$fitted <- fitted(m1)  # Fitted values

# Plotting the residuals against the fitted values
hyp %>%
  mutate(st.resid = resid/sd(resid)) %>% # Making standardized residuals
  ggplot(aes(fitted, st.resid)) + 
  geom_hline(yintercept = 0) +
  geom_point(size = 3, fill = "lightblue", shape = 21) +
  labs(x = "Fitted values",
       y = "Standardized residuals") +
  theme_minimal()
    # All the residuals, except for one, is less than 2.0 SD away from their predicted value. The variance is quite constant.



######## Checking for normal residuals ########
# Checking if the residuals are normal by creating a plot that plot every observed residual against its theoretical position in a normal distribution
hyp %>%
  mutate(st.resid = resid/sd(resid)) %>% 
  ggplot(aes(sample = st.resid)) +
  stat_qq(size = 3, fill = "lightblue", shape = 21) +
  stat_qq_line() +
  labs(x = "Theoretical",
       y = "Sample") +
  theme_minimal()
  # Most of the values follows the straight line. The data is quite normal distributed. 

# Can also show the above using residual plots (four plots).

plot(m1)
par(mfrow = c(2,2)) # Two rows and two coloums


######## Independent observations ########
# The data comes from independent and paired observations, at T1.



######## Do problematic observations matter? ########
# Saw that there was one potentially problematic observation in the residual plot. 
# Labeling the observations in the residual plot to find out what observation is problematic.

hyp %>%
  mutate(st.resid = resid/sd(resid)) %>%
  ggplot(aes(fitted, st.resid, label = SUB_ID)) + 
  geom_hline(yintercept = 0) +
  geom_point(size = 3, fill = "lightblue", shape = 21) +
  geom_label(nudge_x = 20, nudge_y = 0) +
  labs(x = "Fitted values",
       y = "Standardized residuals") +
  theme_minimal()
# Finding that it is subject MRV019 that has the largest residual and and potentially could be a problematic observation

# Checking if the potentially problematic observation MRV019 really is problematic. Doing the model without the observation to see if this changes the conclusion of the analysis.
hyp_reduced <- hypertrophy %>% # Creating a new object without MRV019
  select(SUB_ID, AVG_CSA_T1, Squat_3RM_kg) %>% # Selecting variables of interest
  filter(SUB_ID != "MRV019") %>% # Removing observation MRV019
  filter(!is.na(Squat_3RM_kg), # Removing missing values in Squat_3RM_kg
         !is.na(AVG_CSA_T1))  # Removing missing values in AVG_CSA_T1


# Making a regression model without MRV019
m1_reduced <- lm(Squat_3RM_kg ~ AVG_CSA_T1, data = hyp_reduced)
summary(m1_reduced)
# All the values change a bit

# The delta beta calculates the percentage change in the slope as a consequence of removing the observation with the greatest residual
delta_beta <- (100 * (coef(m1_reduced)[2]/coef(m1)[2] - 1))
# Removing the outlier changes the slope with 61.41295  %. Ergo, quite much.


# We can conclude that the potentially problematic observation **do matter**, and that we can not use all the data points in the analysis. Have to run both the t-test and regression analysis without MRV019. 

## ---------------------------------------------------------------------------------

```

#### Correlation model
There was a `r round(corr, 2)` correlation between average cross-sectional area and 3 RM squat at T1 (95% CI: [`r round(ci1, 3)`, `r round(ci2, 3)`], *t*(`r round(df, 1)`) = `r round(t,3)`, *p* = `r round(p, 2)`).

#### Regression model
```{r, echo=FALSE, results='asis', warning=FALSE, message=FALSE, fig.cap="Figure1: The relationship between muscle mass and strength. Using a regression model with the average cross-sectional area as the predictor, and 3 RM squat as the dependent variable"}
# I use this chunck to print the regression figure

rmodel
rtable

```

When using the regression model and controlling for body mass, there was no relationship between average cross-sectional area and 3 RM squat at T1 (95% CI: [`r round(confint(m1)[2], 3)`, `r round(confint(m1)[5], 3)`], *t*(27) = 0.841, *p* = 0.408). 


## Discussion
#### Correlation model
The correlation estimate (`r round(corr, 2)`) is less than 0.1, which indicates a *trivial* correlation between average cross-sectional area and 3 RM squat at T1 [@Hopkins2009].

As the confidence interval contain 0 (95% CI: [`r round(ci1, 3)`, `r round(ci2, 3)`]), we can not reject the null hypothesis that states that "at T1, there is no correlation between muscle mass measured as average cross-sectional area and muscle strength measured as 3 RM squat, both when and when not controlling for body mass." This is supported by the *p*-value (`r round(p, 2)`).

#### Regression model
We can se from the regression table that for every unit increase in average cross-sectional area, 3 RM squat increases with 0,0028 kg.

As the confidence interval contain 0 (95% CI: [`r round(confint(m1)[2], 3)`, `r round(confint(m1)[5], 3)`]), we can not reject the null hypothesis tat states that "at T1, there is no correlation between muscle mass measured as average cross-sectional area and muscle strength measured as 3 RM squat, both when and when not controlling for body mass." This is supported by the *p*-value (`r round(p, 2)`).

#### Conclusion
The findings from Haun et al. [-@Haun2019] therefore suggests that there is no relationship between average cross-sectional area and 3 RM squat in trained men before starting a training intervention, both when and when not controlling for body mass.


## References

